\documentclass[CT4S-EN-RU]{subfiles}

\begin{document}

\section{\caseENGRUS{What is requested from the student}{ / }{Что потребуется от студента}}

\begin{blockENG}
I will do my best to make clear the value of category theory in science, but I am not a scientist. To that end I am asking for your help in exploring how category theory may be useful in your specific field.
\end{blockENG}

\begin{blockRUS}
Я буду стараться изо всех сил сделать ясной ценность теории категорий в науке, но сам я не ученый. По этой причине я прошу вашей помощи в описании того, как теория категорий может быть полезна в вашей конкретной области. 
\end{blockRUS}

\begin{blockENG}
I also want you to recognize that the value of mathematics is not generally obvious at first. A good student learning a good subject with a good teacher will see something compelling almost immediately, but may not see how it will be useful in real life. This will come later. I hope you will work hard to understand even without yet knowing what its actual value in your life and research will be. Like a student of soccer is encouraged to spend hours juggling the ball when he or she could be practicing penalty shots, it is important to gain facility with the materials you will be using. Doing exercises is imperative for learning mathematics.
\end{blockENG}

\begin{blockRUS}
Я также хочу, чтобы вы поняли, что ценность математики не всегда сразу очевидна. Хороший студент, изучающий хороший предмет у хорошего преподавателя сможет понять нечто впечатляющее практически немедленно, но не сможет так же быстро понять, как это нечто может быть полезно в реальной жизни. Это придет позже. Я надеюсь, вы будете усердно трудиться над предметом, даже еще не понимая, в чем окажется его настоящая ценность в вашей жизни и исследованиях. Учащегося футболу заставляют проводить часы, упражняясь с мячом, вместо того, чтобы попрактиковаться в пенальти - ибо важно заранее обрести умения в избранном деле. Точно так же выполнение упражнений критически важно при обучении математике. 
\end{blockRUS}

\end{document}
